Local PCA Shows How Population Structure Differs Along the Genome

### 2016-06-25 writeup/local_pca_paper.tex
edits in results

The variation in population structure turns out to be almost entirely explained by
several large inversions that are polymorphic in these samples, identified in \citet{corbett2012population}. 
Frequencies of each inversion in the sample are given in Table \ref{tab:inversion_freqs}.
\plr{Add this table.}
We recolored the PCA plots in Figure \ref{fig:pca_by_pop} by the orientation of the inversion for each sample,
obtaining Figure \ref{fig:pca_by_inversion}.
Taking chromosome arm 2L as an example,
the two regions of similar, extreme population structure
shown in green in the first row of Figure \ref{fig:mds_chr2L}
lie directly around the breakpoints of the iversion In(2L)t,
and the PCA plots in the first rows of figure \ref{fig:pca_by_inversion}
shows that population structure here is mostly determined by inversion orientation.
The regions shown in purple lie near the centromere,
and have population structure reflective of two axes of variation,
seen in figure \ref{fig:pca_by_pop},
which correspond roughly to latitude within Africa and to degree of cosmopolitan admixture respectively
(see \citet{pool2015natural} for more details).
\plr{check that reference}
The regions shown in orange mostly lie inside the inversion,
and show population structure that is a mixture between the other two,
as expected due to recombination within the (long) inversion \citep{kirkpatrick2015chromosome}.
\plr{check that reference too}
Similar results are found in other chromosome arms,
\plr{add chromosome X}
except that the situation of Chr3R is a little more complicated,
due to the coexistence of two polymorphic inversions (In(3R)K and In(3R)P) at intermediate frequency.
However, peaks in MDS coordinates along the chromosome show the breakpoints of both,
although we have not highlighted portions of the MDS plot that pick these out.
Two further inversions, In(3L)P and In(3R)Mo, are at very low frequency in these samples, 
and do not significantly influence population structure.

\begin{figure}
    \begin{center}
       \includegraphics{Fig3_pca_plots_color_by_inv_allchr}
    \end{center}
    \caption{
         As in Figure 2, except that samples are colored by orientation of the corresponding polymorphic inversions, 
         In(2L)t, In(2R)NS, In(3L)OK and In(3R)K respectively. 
         (Data from \citet{lack2015drosophila}.
        \label{fig:color_inver}
    }
\end{figure}

@@

As for the \textit{Drosophila} data, we ran our method separately on all 22 human autosomes, 
with results shown in Supplementary Figure XXX.
\plr{Add plots for remaining autosomes to supplement.}
On each, variation in population structure was dominated by a small number of windows
having similar population structure to each other that differed dramatically from the rest of the chromosome.
Such regions seem to often coincide with inversions:
for instance, the eleven windows that are outliers in the first MDS coordinate of chromosome 8 (Figure \ref{fig:mds_human}b) 
coincide with the position of known polymorphic inversions on 8p23. 
Similar results are found in other chromosomes that have known inversions,
shown in Figure \ref{fig:mds_human}.
PCA plots of these windows show a characteristic trimodal shape,
presumably distinguishing samples having each of the three diploid genotypes for each inversion orientation
(although we do not have data on orientation status).
PCA has been previously proposed as a method to computationally identify inversions \citep{ma2012investigation},
but understanding how well the method can distinguish inversions from regions of low recombination
is beyond the scope of this paper.

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{Fig4_POPRES_Together_MDS_plot_chr8_15_17}
    \end{center}
    \caption{
         Variation in structure between windows on human chromosomes 8, 15, and 17. 
         Each point in each plot represents a window.
         The first column shows the MDS visualization of relationships between windows;
         the second and third columns show the two MDS coordinates of each window against its position (midpoint) along the chromosome. 
         Rows, from top to bottom show chromosomes 8, 15, and 17. 
         The vertical red lines show the breakpoints of known inversions from \citep{antonacci2009characterization},
         whose values are listed in \plr{add these in a table}.
        \label{fig:mds_human}
    }
\end{figure}

@@

Unlike the other two species,
the method applied separately on all eight chromosomes of \textit{Medicago truncatula} 
showed similar patterns of gradual change in population structure across each chromosome,
with no indications of chromosome-specific patterns.
This consistency suggests that the factor driving the population structure for each chromosome is the same,
as might be caused by varying strengths of linked selection.
To verify that variation in population structure is shared across chromosomes,
we applied the method to all chromosomes together.
Results for chromosome 3 are shown in Figures \ref{fig:mds12_medicago} and \ref{fig:pca_medicago}
(other chromosomes are similar).

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{Fig6_Together_MDS_plot_chr3_final}
    \end{center}
    \caption{
     MDS visualization for Medicago chromosome 3. Each point in the plot represents a window.
       \label{fig:mds12_medicago}
    }
\end{figure}

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{Fig7_pca_plots_for_Medicago_chr3_3peaks_byMDS}
    \end{center}
    \caption{
        PCA plots for the sets of genomic windows colored (A) green, (B) orange, and (C) purple in Figure \ref{fig:mds12_medicago}. 
        Each point corresponds to a sample, colored by country of origin.
        \label{fig:pca_medicago}
    }
\end{figure}

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=0.82\textwidth]{FigS_pca_plots_for_Medicago_allchr_3peaks_byMDS}
    \end{center}
    \caption{
        PCA plots for all 8 chromosomes' 3 peaks in \textit{Medicago truncatula}.
        \label{fig:pca_peaks_medicago_allchr}
    }
\end{figure}

>>>>

The variation in population structure turns out to be almost entirely explained by
several large inversions that are polymorphic in these samples, identified in \citet{corbett2012population}. 
% Frequencies of each inversion in the sample are given in Table \ref{tab:inversion_freqs}.
We recolored the PCA plots in Figure \ref{fig:pca_by_pop} by the orientation of the inversion for each sample,
obtaining Figure \ref{fig:pca_by_inversion}.
Taking chromosome arm 2L as an example,
the two regions of similar, extreme population structure
shown in green in the first row of Figure \ref{fig:mds_chr2L}
lie directly around the breakpoints of the iversion In(2L)t,
and the PCA plots in the first rows of figure \ref{fig:pca_by_inversion}
shows that population structure here is mostly determined by inversion orientation.
The regions shown in purple lie near the centromere,
and have population structure reflective of two axes of variation,
seen in figure \ref{fig:pca_by_pop},
which correspond roughly to latitude within Africa and to degree of cosmopolitan admixture respectively
(see \citet{pool2015natural} for more details).
\plr{check that reference}
The regions shown in orange mostly lie inside the inversion,
and show population structure that is a mixture between the other two,
as expected due to recombination within the (long) inversion \citep{kirkpatrick2015chromosome}.
\plr{check that reference too}
Similar results are found in other chromosome arms,
\plr{add chromosome X}
except that the situation of Chr3R is a little more complicated,
due to the coexistence of two polymorphic inversions (In(3R)K and In(3R)P) at intermediate frequency.
However, peaks in MDS coordinates along the chromosome show the breakpoints of both,
although we have not highlighted portions of the MDS plot that pick these out.
Two further inversions, In(3L)P and In(3R)Mo, are at very low frequency in these samples, 
and do not significantly influence population structure.

\begin{figure}
    \begin{center}
       \includegraphics{Fig3_pca_plots_color_by_inv_allchr}
    \end{center}
    \caption{
         As in Figure 2, except that samples are colored by orientation of the corresponding polymorphic inversions, 
         In(2L)t, In(2R)NS, In(3L)OK and In(3R)K respectively. 
         (Data from \citet{lack2015drosophila}.
        \label{fig:pca_by_inversion}
    }
\end{figure}

@@

As for the \textit{Drosophila} data, we ran our method separately on all 22 human autosomes, 
with results shown in Supplementary Figure XXX.
\plr{Add plots for remaining autosomes to supplement.}
On each, variation in population structure was dominated by a small number of windows
having similar population structure to each other that differed dramatically from the rest of the chromosome.
These may be primarily inversions:
for instance, the eleven windows that are outliers in the first MDS coordinate of chromosome 8 (Figure \ref{fig:mds_human}b) 
coincide with the position of a particularly large, polymorphic inversion on 8p23. 
More generally, we see signals coincident with 
three of the six large polymorphic inversions described in \citet{antonacci2009characterization},
shown in Figure \ref{fig:mds_human}.
PCA plots of these windows show a characteristic trimodal shape,
presumably distinguishing samples having each of the three diploid genotypes for each inversion orientation
(although we do not have data on orientation status),
a signal that has been proposed as a method to identify inversions \citep{ma2012investigation}.
However, distinguishing this hypothesis from others,
such as regions of low recombination rate,
would require additional data.

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{Fig4_POPRES_Together_MDS_plot_chr8_15_17}
    \end{center}
    \caption{
         Variation in structure between windows on human chromosomes 8, 15, and 17. 
         Each point in each plot represents a window.
         The first column shows the MDS visualization of relationships between windows;
         the second and third columns show the two MDS coordinates of each window against its position (midpoint) along the chromosome. 
         Rows, from top to bottom show chromosomes 8, 15, and 17. 
         The vertical red lines show the breakpoints of known inversions from \citep{antonacci2009characterization}.
        \label{fig:mds_human}
    }
\end{figure}

@@

Unlike the other two species,
the method applied separately on all eight chromosomes of \textit{Medicago truncatula} 
showed similar patterns of gradual change in population structure across each chromosome,
with no indications of chromosome-specific patterns.
This consistency suggests that the factor driving the population structure for each chromosome is the same,
as might be caused by varying strengths of linked selection.
To verify that variation in population structure is shared across chromosomes,
we applied the method to all chromosomes together.
Results for chromosome 3 are shown in Figures \ref{fig:mds12_medicago} and \ref{fig:pca_medicago},
and other chromosomes are similar:
the two types of population structure with higher gene density (colored green and purple in Figure \ref{fig:mds12_medicago}
are nearly identical across all chromosomes,
but population structure associated with low gene density (colored orange)
shows more chromsome-specific patterns (see Supplemental Figure \ref{fig:pca_drosophila_allchr}).

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{Fig6_Together_MDS_plot_chr3_final}
    \end{center}
    \caption{
     MDS visualization for Medicago chromosome 3. Each point in the plot represents a window.
     Plots for remaining chromosomes are shown in Supplemental Figure \ref{fig:mds_medicago_allchr}.
       \label{fig:mds12_medicago}
    }
\end{figure}

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{Fig7_pca_plots_for_Medicago_chr3_3peaks_byMDS}
    \end{center}
    \caption{
        PCA plots for the sets of genomic windows colored (A) green, (B) orange, and (C) purple in Figure \ref{fig:mds12_medicago}. 
        Each point corresponds to a sample, colored by country of origin.
        Plots for remaining chromosomes are shown in Supplemental Figure \ref{fig:pca_peaks_medicago_allchr}.
        \label{fig:pca_medicago}
    }
\end{figure}

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=0.82\textwidth]{FigS_pca_plots_for_Medicago_allchr_3peaks_byMDS}
    \end{center}
    \caption{
        PCA plots for all 8 chromosomes' 3 peaks in \textit{Medicago truncatula}.
        (A) green, (B) orange, and (C) purple.
        \label{fig:pca_peaks_medicago_allchr}
    }
\end{figure}
